
---
title: "PHI 420/555 - Race Causality Discrimination"
subtitle: "Fall 23"
author: "[Marcello Di Bello](https://www.marcellodibello.com/)  - mdibello@asu.edu"
output:
    rmdformats::html_clean:
    fig_width: 6
    fig_height: 6
    highlight: kate
    thumbnails: true
    lightbox: true
    gallery: true
    toc_depth: 2
    toc: TRUE
---    






    






<script>
   $(document).ready(function() {
     $head = $('#header');
     $head.prepend('<img src=\"algo-fair-logo.png\" style=\"float: right;width: 300px;height: 220px;\"/>')
   });
</script>



```{r knitr_init, echo=FALSE, results="asis", cache=FALSE}
library(knitr)
library(rmdformats)
## Global options
options(max.print = "75")
opts_chunk$set(echo = FALSE,
	             cache = FALSE,
               prompt = FALSE,
               tidy = TRUE,
               comment = NA,
               message = FALSE,
               warning = FALSE)
opts_knit$set(width = 75)
```



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


<style type="text/css">

body{ /* Normal  */
      font-size: 20px;
      font-family:'Avenir Next';
  }

</style>



<!--- 
ADDED STYLES/ FONT SIZSES
<style type="text/css">
body{ /* Normal  */
      font-size: 18px;
  }
  
 
td {  /* Table  */
  font-size: 8px;
}
h1.title {
  font-size: 38px;
  color: DarkRed;
}
h1 { /* Header 1 */
  font-size: 28px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>
-->


Tue	10:30-1:20 PM in Tempe, [COOR](https://www.asu.edu/map/interactive/?psCode=COOR) ~~3301~~ 3323  

<!-- ~~3301~~ 3323   --->


Office hours: Tuesday after class or by appointment

# Topic

Claims of racial discrimination and structural racism are ubiquitous today. On one hand, studies in the social sciences show that, after controlling for variables other than race, racial disparities still persist in incarceration, health outcomes and wealth. Thus, race appears to be a key variable for explaining racial disparities. On the other hand, many point out that, again after controlling for variables other than race, the same racial disparities significantly decrease. Thus, they think that the explanatory role of race has been exaggerated. This debate leaves out a crucial question, however. What causal and explanatory role (if any) can race play in the first place? If, as many hold, race is a social construct emerging from the fabric of social interactions, does it make sense to isolate it and control for other variables? 

The course aims to untangle the assumptions about the causality of race that inform the ongoing debate about racial discrimination and structural racism. Readings will be drawn from the literature in the social sciences on racial inequalities, the statistical and philosophical literature on causality, and the philosophical literature on theories of race. 


# Objectives

This course is meant for advanced undergraduates and graduate students. 
You will become familiar with current debates in philosophy and the social sciences 
about the nature and study of causality, with a focus on 
questions about race and discrimination. You will sharpen your critical thinking 
skills, in reading  and  writing, for the analysis of theoretical concepts and their application to current social and political questions. You will read academic papers in philosophy 
and the social sciences, and develop an appreciation for their distinctive contributions. 

# Course Materials

Readings and other course materials are available 
on this website. For readings covered 
by copyright, please check the Canvas page of this course.  
Course materials are divided into "essential" and "additional". 
You are only expected to study the essential ones, but I 
recommend that you have a look at the 
additional materials for at least one 
or two weeks.


# Requirements


## Participation

Please attend class regularly and participate 
(**10% of your grade**). This is a "seminar style" course. You are 
expected to take an active role in the discussions. Please study the assigned 
materials **before class** and be ready to discuss them. 


## Assignments

In addition, 
you should write ten 
Pass/Fail précis as well as three graded essays or 
a research paper (**90% of your grade**).

### Pass/fail 

Every week please write a **one-page précis** of one of the papers assigned for that week. 
The précis should describe: (a) topic of the paper; 
(b) main thesis (or main theses, if there are more than one); 
(c) supporting arguments; (d) objections to these arguments, complications or 
difficulties that the author considers (if any). Submit 
your précis each week through Canvas **before the beginning of class**. 
You should receive a PASS in **at least ten précis**, or else a full letter grade 
will be subtracted from your final grade: A would become B; B would become C; etc.


### Essays 

There will be three main graded assignments 
for this course, **5 pages each**. 

**1**. Write a **critical essay** that discusses the manipulationist account of causation.
Your essay should begin with a careful theoretical summary of the key ideas in the manipulationist account (1-2 pages). After that, you should apply the account to a specific example in the social sciences and show how the account can be useful in modeling causation (1-2 pages). Finally, you should raise problems and difficulties for the account (1-2- pages). Please draw from the course readings as needed and demonstrate thoughful engagement with them. 

**2**. After doing independent research, collect and summarize a **case study** that illustrates the challenges of
investigating the causal role of race. The first part of your essay (1-2 pages) should describe your case study. Topics for your case study could include, but are not limited to, health care, criminal justice, education, hiring.  The second part (3-4 pages) should draw from the readings on the philosophy of race. 

**3**. Write a **philosophical argumentative essay** that examines the topic "Philosophical Challenges to the Study of Racial Discrimination". In examining this topic, you should draw from the course readings on racial discrimination. You may (but need not) build on the case study you discussed in your second essay.  Your essay can be primarily negative, e.g., you argue that the challenges are insurmountable. Your essay can also be more positive, e.g., you describe the challenges and propose ways to address them. 


### Research paper 

If you are a grad student or advanced undergrad with research experience, 
you may combine the three 5 page essays into one 
**15-20 page** research paper. The research paper should engage closely 
with a subset of the course materials including the additional ones. 
It is neither necessary nor recommended that you use materials outside those already 
included in the course materials. 

*Please come talk to me before you start working on the research paper.* 



# Handouts 

[Week 1](https://www.marcellodibello.com/race-causality-discrimination/handout/introduction.html) |
[Week 2](https://www.marcellodibello.com/race-causality-discrimination/handout/causation-manipulation.html) |
[Week 3](https://www.marcellodibello.com/race-causality-discrimination/handout/causation-manipulation-Woodward.html) |
[Week 4](https://www.marcellodibello.com/race-causality-discrimination/handout/race-cause.html) |
[Week 5](https://www.marcellodibello.com/race-causality-discrimination/handout/audit-studies.html) |
[Week 7](https://www.marcellodibello.com/race-causality-discrimination/handout/race-construction.html) |
[Week 9](https://www.marcellodibello.com/race-causality-discrimination/handout/race-real.html) |
[Week 10](https://www.marcellodibello.com/race-causality-discrimination/handout/discrimination.html) |
[Week 11](https://www.marcellodibello.com/race-causality-discrimination/handout/test-discrimination.html)


# Schedule 

## 8/22

**Week 1: Introduction**

Racial disparities are well-known and widely studied. However, it is more difficult to study 
the *causal* effects of race on outcomes. For example, in a 2017  [study](https://scholar.harvard.edu/sites/scholar.harvard.edu/files/fryer/files/empirical_analysis_tables_figures.pdf) Harvard economist [Ronald 
Fryer](https://scholar.harvard.edu/fryer/home) argued that available data show no racial bias against black men in police killings. In fact, Fryer's study showed that 
black civilians are *less* likely to be killed by the police compared to similarly situated white civilians.
Many rejected this conclusion despite the study's meticulous reliance on data and rigorous statistical modeling.
What makes it so difficult to study the causal effects of race? Are data-driven social sciences inadequate for this task?

*Essential readings*

 
- [Handout - Week 1](https://www.marcellodibello.com/race-causality-discrimination/handout/introduction.html) 


- Hu (2021), [Race, Policing, and the Limits of Social Science](https://www.bostonreview.net/articles/race-policing-and-the-limits-of-social-science/)

- Hu (2019), [Disparate Causes, Part I](https://www.phenomenalworld.org/analysis/disparate-causes-i/)

*Additional readings and materials*

- Fryer (2017), [An Empirical Analysis of Racial Differences in Police Use of Force](https://scholar.harvard.edu/sites/scholar.harvard.edu/files/fryer/files/empirical_analysis_tables_figures.pdf) 
(see also the author's [response](https://www.journals.uchicago.edu/doi/abs/10.1086/710977) to critics)

## 8/29

**Week 2: Causation as Manipulation in Statistics**

How should we understand the nature of causality and how should we study causality? A prominent account in both statistics and philosophy is the so-called *manipulationist account*. This week and the next will be devoted to understanding this manipulationist account. (Since the main reading 
requires an understanding of conditional probabilities, you can check out this 
[video](https://www.youtube.com/watch?v=_IgyaD7vOOA) if you are unfamiliar with this notion.)


*Essential readings*

- [Handout - Week 2](https://www.marcellodibello.com/race-causality-discrimination/handout/causation-manipulation.html) 


- Holland (1986), [Statistics and Causal Inference](https://www.marcellodibello.com/race-causality-discrimination/readings/Holland-StatisticsCausalInference-1986.pdf)

*Additional readings and materials*

- Pearl (2003), [Statistics and Causal Inference: A Review](https://citeseerx.ist.psu.edu/document?repid=rep1&type=pdf&doi=211680a75afaa0a4130a47530af554ea0de75629)

##  9/5


**Week 3: Causation as Manipulation in Philosophy**

We continue on the same topic as last week, focusing 
on work by philosophers. 


*Essential readings*

- [Handout - Week 3](https://www.marcellodibello.com/race-causality-discrimination/handout/causation-manipulation-Woodward.html)

- Woodward (2003), Causation and Manipulation, Chapter 1 and 2 of the book *Making Things Happens*, 
Oxford University Press - *check Canvas*
 

*Additional readings and materials*

- Hall (2007), [Structural Equations and Causation](https://www.marcellodibello.com/race-causality-discrimination/readings/Hall-StructuralEquationsCausation-2007.pdf)



## 9/12


**Week 4: Race and Manipulation**

The manipulationist account of causation faces several challenges when it comes to investigating the causal role of social categories, in particular, race. The main challenge is that 
race cannot be a cause insofar as race cannot be manipulated. 
And yet, even if race is not a variable that can be manipulated, it could still play a causal role, just like 
class or gender can have a causal role. The slogan 
"no causation without manipulation", then, seems overly restrictive. 


*Essential readings*

- [Handout - Week 4](https://www.marcellodibello.com/race-causality-discrimination/handout/race-cause.html)



- Holland (2003), [Causation and Race](https://www.marcellodibello.com/race-causality-discrimination/readings/Holland-Causation-Race-2003.pdf)

- Marcellesi (2013), [Is Race a Cause?](https://www.marcellodibello.com/race-causality-discrimination/readings/Marcellesi-RaceCause-2013.pdf)


*Additional readings and materials*

- Krieger (2014), [Discrimination and Health Inequities.](https://www.marcellodibello.com/race-causality-discrimination/readings/Krieger-Discrimination-HealthInequities-2014.pdf) 
(This article reviews the literature on the causal effects of race, as found in studies in epidemiology and health. It illustrates how the causal effects of race can be studied empirically without necessarily presupposing that race is a variable to be manipulated.)



## 9/19

**First essay due**


**Week 5: Audit Studies**

How do social scientists respond to these challenges? We will look at a few examples of studies in sociology and economics that make causal claims about race. In these studies, race (or to be more precise, a variable that is an indicator of race) is still manipulated. Is the manipulationist account, then, secure from objections when it comes to social categories such as race?

**Q&A with Lily Hu**

The last part of the class will consist in a Q&A session with [Lily Hu](https://scholar.google.com/citations?user=oWI7EukAAAAJ&hl=en), a philosopher at Yale interested in how the social sciences study the causal effects of race. Hu is the author of one of the readings assigned for today.

*Essential readings*

- [Handout - Week 5](https://www.marcellodibello.com/race-causality-discrimination/handout/audit-studies.html)


- Hu (2002), Interventionist Social Causes: The Case of Race and Sex, Chapter 2 of [*Causation in the Social World*](https://www.marcellodibello.com/race-causality-discrimination/readings/Hu-CausalitySocial2022.pdf)

- Pager (2007),  [The Use of Field Experiments for Studies of Employment Discrimination](https://scholar.harvard.edu/sites/scholar.harvard.edu/files/pager/files/annals_pager.pdf) 

*Additional readings and materials*


- Bertrand and Mullainathan (2004), [Are Emily and Greg More Employable than Lakisha and Jamal?](https://cos.gatech.edu/facultyres/Diversity_Studies/Bertrand_LakishaJamal.pdf) 


- Greiner and Rubin (2010), [Causal Effects of Perceived Immutable Characteristics](https://www.marcellodibello.com/race-causality-discrimination/readings/Greiner-Rubin-Cause-Immutable-2010.pdf)



## 9/26


**Week 6: Race as a Bundle of Sticks**

Today we look at the theory of race as a "bundle of sticks" and the possibility 
of manipulating other characteristics that may then be used 
as indicators of the causal role of race.


*Essential readings*

- Sen and Wasow (2016), [Race as a Bundle of Sticks](https://scholar.harvard.edu/files/msen/files/race_causality.pdf)

- Walton and Cohen (2011), [A Brief Social-belonging Intervention Improves Academic and Health Outcomes of Minority Students](https://www.marcellodibello.com/race-causality-discrimination/readings/Walton-BriefSocialBelongingIntervention-2011.pdf)

*Additional readings and materials*

- VanderWeele and Robinson (2014), [On Causal Interpretation of Race in Regressions Adjusting for Confounding and Mediating Variables](https://www.marcellodibello.com/race-causality-discrimination/readings/VanderWeeleRobinson-RaceRegression2024.pdf)



## 10/3


**Week 7: Race as a Social Construct**


It is time to examine philosophical accounts of race. Admittedly, we cannot talk about whether race is a cause without asking, what is race? So we ask this question now. We'll start with social constructionist views of race. 

**Q&A with Sally Haslanger**

The first part of the class will consist in a Q&A session with Sally Haslanger, a philosopher at Massachusetts Institute of Technology interested in social practices, social structure, structural explanation with an emphasis on the materiality of social practices and the role of ideologyn. Sally is the author of one of the readings assigned for today.


*Essential readings*


- [Handout - Week 7](https://www.marcellodibello.com/race-causality-discrimination/handout/race-construction.html)


- Haslanger (2019), Tracing the Sociopolitical Reality of Race, in [What is Race? Four Philosophical Views](https://academic.oup.com/book/34975), Oxford University Press

- Jeffers (2019), Cultural Contructionism, in [What is Race? Four Philosophical Views](https://academic.oup.com/book/34975), Oxford University Press





## 10/10


**Week 8: Fall Break, No class**



## 10/17

**Week 9: Race as Real or Unreal**

This week we will look at accounts of race as biologically 
real and as an illusion. 

*Essential readings*

- [Handout - Week 9](https://www.marcellodibello.com/race-causality-discrimination/handout/race-real)


- Spencer (2019), How to Be a Biological Racial Realist, in [What is Race? Four Philosophical Views](https://academic.oup.com/book/34975), Oxford University Press

- Glasgow (2019), Is Race an Illusion or a (Very) Basic Reality?, in [What is Race? Four Philosophical Views](https://academic.oup.com/book/34975), Oxford University Press


*Additional readings and materials*

-  Appiah (1985), [The Uncompleted Argument: Du Bois and the Illusion of Race](https://blogs.law.columbia.edu/revolution1313/files/2022/01/Anthony-Appiah-The-Uncompleted-Argument-Du-Bois-and-the-Illusion-of-Race.pdf)

- Mallon (2006), [‘Race': Normative, Not Metaphysical or Semantic](https://www.marcellodibello.com/race-causality-discrimination/readings/Mallon-RaceNormativeMetaphysical-2006.pdf)

## 10/24


**Second essay due**



**Week 10: Measuring Racial Discrimination**

Having clarified what we could mean by race, let's return to our original question:  how can we study the causal effects of race? We will focus in particular on racial discrimination, one crucial way in which race can have causal effects. 
What is discrimination and how is it measured?


*Essential readings*


- [Handout - Week 10](https://www.marcellodibello.com/race-causality-discrimination/handout/discrimination.html)

- (2004), [Causal Inference and the Assessment of Racial Discrimination](https://nap.nationalacademies.org/read/10887/chapter/9), Chapter 5 of *Measuring Racial Discrimination*, The National Academies Press


-  Barocas,  Hardt and Narayanan (2023)
[Causality](https://fairmlbook.org/causal.html), especially section “Counterfactuals in the law”, Chapter 5 of [Fairness and Machine Learning](https://fairmlbook.org/)




## 10/31


**Week 11: Testing for Discrimination**


Now, if race is socially constructed, how can it be subject to statistical analysis by manipulating it as a variable or controlling for other variables? What are the pitfalls of this standard way of studying discrimination? 

**Q&A with Naftali Weinberger**

The first part of the class will consist in a Q&A session with [Naftali Weinberger](https://sites.google.com/site/naftaliweinberger/), a philosopher at the Munich Center for Mathematical Philosophy, Germany, interested in causally modeling and causal issues related to discrimination. 
Naftali is the author of one of the readings assigned for today.


*Essential readings*

- [Handout - Week 11](https://www.marcellodibello.com/race-causality-discrimination/handout/test-discrimination.html)


- Kohler-Hausmann (2019), [Eddie Murphy and the Dangers of Counterfactual Causal Thinking about Detecting Racial Discrimination](https://scholarlycommons.law.northwestern.edu/cgi/viewcontent.cgi?article=1374&context=nulr)

- Naftali Weinberger (2023), [Signal Manipulation and the Causal Analysis of Racial Discrimination](https://www.marcellodibello.com/race-causality-discrimination/readings/Signal-Manipulation-Causal-Race-2023.pdf)


*Additional readings and materials*


- Hu (2023), [What Is ‘Race’ in Algorithmic Discrimination on the Basis of Race?](https://scholar.harvard.edu/sites/scholar.harvard.edu/files/lilyhu/files/what_is_race.pdf)


## 11/07


**Week 12: Race and Social Structure**

Instead of thinking of race as a single variable, race can be thought of
as embedded in a complex social structure. This week we examine an idea that 
is often mentioned, even in the mainstream media, namely "structural injustice". 


*Essential readings*

- Haslanger (2012), [Oppression: Racial and Other](), chapter 11 of the book *Resisting Reality*, Oxford University Press - *check Canvas*

- Haslanger (2015), [What Is a (Social) Structural Explanation?]() 

*Additional readings and materials*

- Jackson and Pettit (1992), [Structural Explanation in Social Theory](), in  Charles and Lennon (eds), Reduction, explanation and realism, Oxford University Press



## 11/14


**Week 13: Social Structural Explanation**


To what extent can a manipulationist account of causality model the causal role of race in a structural sense? Some think this framework can capture the structural picture. How do such accounts model the causal (structural?) role of race? 

**Q&A with Lauren Ross**

The first part of the class will consist in a Q&A session with [Lauren Ross](https://sites.socsci.uci.edu/~rossl/), a philosopher at University of California, Irvine interested in causal reasoning and the nature of explanation. Lauren is the author of one of the readings assigned for today.


*Essential readings*


- Ross (2023), [What Is Social Structural Explanation? A Causal Account](https://www.marcellodibello.com/race-causality-discrimination/readings/Ross-SsocialStructExplNous-2023 .pdf)


- Malinsky (2018), [Intervening on structure](https://www.marcellodibello.com/race-causality-discrimination/readings/Malinski-InterveningStructure-2018.pdf)


*Additional readings and materials*

-  Bright, Malinsky, and  Thompson (2016), [Causally Interpreting Intersectionality Theory](https://www.liamkofibright.com/uploads/4/8/9/8/48985425/causally_interpreting_intersectionality_theory_final.pdf)


## 11/21

**Week 14:  Workshop**


Students will workshop their research papers by giving short presentations accompanied by either a handout or slides and followed by Q&A. Each presenter will have 15 minutes to be split between the presentation itself and Q&A at the discretion of the presenter. The list of presenters is below.  

Workshop details TBA. 



## 11/28

**Third essay or research paper due**

**Week 15: Conclusion**

What have we learned about race, causality and discrimination?
What questions are still left open and 
what lines of inquiry seemed more promising?



# Title IX 

Title IX is a federal law that provides that no person be excluded on the basis of sex from participation in, be denied benefits of, or be subjected to discrimination under any education program or activity.  Both Title IX and university policy make clear that sexual violence and harassment based on sex is prohibited.  An individual who believes they have been subjected to sexual violence or harassed on the basis of sex can seek support, including counseling and academic support, from the university.  If you or someone you know has been harassed on the basis of sex or sexually assaulted, you can find information and resources at https://sexualviolenceprevention.asu.edu/faqs. 

 

As a mandated reporter, I am obligated to report any information I become aware of regarding alleged acts of sexual discrimination, including sexual violence and dating violence.  ASU Counseling Services, https://eoss.asu.edu/counseling is available if you wish to discuss any concerns confidentially and privately. ASU online students may access 360 Life Services, https://goto.asuonline.asu.edu/success/online-resources.html.
