
---
title: "Race Causality Discrimination"
subtitle: "Fall 23"
author: "[Marcello Di Bello](https://www.marcellodibello.com/)  - mdibello@asu.edu"
output:
    rmdformats::html_clean:
    fig_width: 6
    fig_height: 6
    highlight: kate
    thumbnails: true
    lightbox: true
    gallery: true
    toc_depth: 2
    toc: TRUE
---    






    






<script>
   $(document).ready(function() {
     $head = $('#header');
     $head.prepend('<img src=\"algo-fair-logo.png\" style=\"float: right;width: 300px;height: 220px;\"/>')
   });
</script>



```{r knitr_init, echo=FALSE, results="asis", cache=FALSE}
library(knitr)
library(rmdformats)
## Global options
options(max.print = "75")
opts_chunk$set(echo = FALSE,
	             cache = FALSE,
               prompt = FALSE,
               tidy = TRUE,
               comment = NA,
               message = FALSE,
               warning = FALSE)
opts_knit$set(width = 75)
```



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


<style type="text/css">

body{ /* Normal  */
      font-size: 20px;
      font-family:'Avenir Next';
  }

</style>



<!--- 
ADDED STYLES/ FONT SIZSES
<style type="text/css">
body{ /* Normal  */
      font-size: 18px;
  }
  
 
td {  /* Table  */
  font-size: 8px;
}
h1.title {
  font-size: 38px;
  color: DarkRed;
}
h1 { /* Header 1 */
  font-size: 28px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>
-->


Th	1:30-4:15 PM in Tempe, [COOR](https://www.asu.edu/map/interactive/?psCode=COOR) ~~3301~~ 3323   

Office hours: Thursday after class or by appointment

# Topic

This seminar examines 



# Objectives

This course is meant for advanced undergraduates and graduate students. 
You will become familiar with current debates in philosophy and the social sciences 
about the nature and study of causality in the social realm, with a focus on 
questions about race and discrimination. You will sharpen your critical thinking 
skills, in reading  and  writing, for the analysis of theoretical concepts and their application to current 
social and political questions. You will read academic papers in philosophy 
and the social sciences, and develop an appreciation for their distinctive contribution. 

# Course Materials

Readings and other course materials are available 
on this website. For readings covered 
by copyright, please check the Canvas page for this course.  
Course materials are divided into "essential" and "additional". 
You are only expected to study the essential ones, but I 
recommend that you have a look at the 
additional materials for at least one 
or two weeks.


# Requirements


## Participation

Please attend class regularly and participate 
(**10% of our grade**). This is a "seminar style" course. You are 
expected to take an active role in the discussions. Please study the assigned 
materials **before class** and be ready to discuss them. 


## Assignments

In addition, 
you should write ten 
Pass/Fail précis as well as three graded essays or 
a research paper (**90% of your grade**).

### Pass/fail 

Every week please write a **one-page précis** of one of the papers assigned for that week. 
The précis should describe: (a) topic of the paper; 
(b) main thesis (or main theses, if there are more than one); 
(c) supporting arguments; (d) objections to these arguments, complications or 
difficulties that the author considers (if any). Submit 
your précis each week through Canvas **before the beginning of class**. 
You should receive a PASS in **at least ten précis**, or else a full letter grade 
will be subtracted from your final grade: A would become B; B would become C; etc.


### Essays 

There will be three main graded assignments 
for this course, **5 pages each**. 

**1**. Write a **critical response** to one of the philosophy papers you read. 
Your response should contain a careful summary of the the specific thesis and argument of the paper you want to criticize (1-2 pages). After that, you should explain, in great detail and with painstaking precision, why you disagree with the paper's thesis and argument (3-4 pages). Please select a paper you disagree with. If you agree with everything you read, please come talk to me.


**2**. After doing independent research, collect and summarize a **case study** that illustrates the idea of the common ground (or lack thereof). The first part of your essay (1-2 pages) should describe your case study. Topics for your case study could include, but are not limited to, debates about abortion, vaccine, race, climate change. The second part (3-4 pages) should use some of the course readings to analyze your case study from the point of view of philosophy of language, logic or epistemology. 

**3**. Write a **philosophical argumentative essay** that attempts to answer the question "What Is the Common Ground and Why Does it Matter?" In answering the first part of the question, I recommend that you focus on a specific notion of the common ground, as is discussed in philosophy of language, logic, epistemology, social and political philosophy, or post-colonial theory. In answering the second part of the question, I recommend that you draw from current trends in society and politics. You may (but need not to) build on the case study you discussed in your second graded assignment.  You are encouraged to think critically about the common ground, highlighting its positive as well as its negative implications. 


### Research paper 

If you are a grad student or advanced undergrad with research experience, 
you may combine the three 5 page essays into one 
**15-20 page** research paper. The research paper should engage closely 
with a subset of the course materials including the additional ones. 
It is neither necessary nor recommended that you use materials outside those already 
included in the course materials. 

*Please come talk to me before you start working on the research paper.* 



# Handouts 

[Week 1](https://www.marcellodibello.com/race-causality-discrimination/handout/introduction.html) | 



# Schedule 

## 8/22

**Week 1: Introduction**

*Essential readings*

- [Week 1 - handout](https://www.marcellodibello.com/race-causality-discrimination/handout/introduction.html) 




## 8/29

**Week 2: Causation as Manipulation in Statistics**

How should we understand the nature of causality and how should we study causality? A prominent account in both statistics and philosophy is the so-called *manipulationist account*. This account aims to guide both our philosophical theorizing about the nature of causality, as well as the emprical study of causal effects. 
This week and the next will be devoted to understanding this manipulationist account. 


*Essential readings*

- [Week 2 - handout](https://www.marcellodibello.com/race-causality-discrimination/handout/.html)  

- Holland (1986), [Statistics and Causal Inference](http://people.umass.edu/~stanek/pdffiles/causal-holland.pdf)

*Additional readings and materials*



##  9/5


**Week 3: Causation as Manipulation in Philosophy**


*Essential readings*

- [Week 3 - handout](https://www.marcellodibello.com/race-causality-discrimination/handout/.html)  

- Woodward (2003), Causation and Manipulation, Chapter 2 of Woodward's book *Making Things Happens*, 
Oxford University Press
 

*Additional readings and materials*

- Hall (2007), [Structural Equations and Causation](https://www.marcellodibello.com/race-causality-discrimination/readings/Hall-StructuralEquationsCausation-2007.pdf)



## 9/12


**Week 4: Race and Manipulation: no!**

We will now look at the challenges that this manipulationist account of causation faces in investigating the causal role of social categories, in particular, race. A number of authors think that race cannot be a cause. If, as the slogan goes, "no causation without manipulation", then race cannot be a cause insofar as race cannot be manipulated. 
But isn't this view too radical? Perhaps, race is not 
a variable that can be manipulated, but it still has a causal role, just like 
class or gender can well have a causal role.  


*Essential readings*

- [Week 4 - handout](https://www.marcellodibello.com/race-causality-discrimination/handout/.html)  

- Holland (2003), [Causation and Race](https://www.marcellodibello.com/race-causality-discrimination/readings/Holland-Causation-Race-2003.pdf)

- Marcellesi (2013), Is Race a Cause?


*Additional readings and materials*

- Krieger (2014), [Discrimination and Health Inequities.](https://www.marcellodibello.com/race-causality-discrimination/readings/Krieger-Discrimination-HealthInequities-2014.pdf) 
(This article reviews the literature on the causal effects of race, as found in studies in epidemiology and health. It illustrates how the causal effects of race can be studied empirically without necessarily presupposing that race is a variable to be manipulated.)



## 9/19

**Week 5: Race and Manipulation: yes!**

How do social scientists respond to these challenges? We will look at a few examples of studies in sociology and economics that make causal claims about race. In these studies, race (or a signal for race) is still manipulated. Is the manipulationist account, then, secure from objections when it comes to social categories such as race?


*Essential readings*

- [Week 5 - handout](https://www.marcellodibello.com/race-causality-discrimination/handout/.html)  

- Pager (2007),  [The use of field experiments for studies of employment discrimination: contributions, critiques, and directions for the future](https://scholar.harvard.edu/sites/scholar.harvard.edu/files/pager/files/annals_pager.pdf) 

Greiner and Rubin (2010), [Causal effects of perceived immutable characteristics](https://www.marcellodibello.com/race-causality-discrimination/readings/Greiner-Rubin-Cause-Immutable-2010.pdf)


*Additional readings and materials*

- Bertrand and Mullainathan (2004), [Are Emily and Greg more employable than Lakisha and Jamal?](https://cos.gatech.edu/facultyres/Diversity_Studies/Bertrand_LakishaJamal.pdf) 



## 9/26


**Week 6: Race as Bundle of Sticks**

We look at the theory of race as a bundle of sticks and the possibility 
of manipulating other characteristics that may then be used 
as indicators of the causal role of race.


*Essential readings*

- [Week 6 - handout](https://www.marcellodibello.com/race-causality-discrimination/handout/.html)  

- Sen and Wasow (2016), [Race as a bundle of sticks: Designs that estimate effects of seemingly immutable characteristics](https://scholar.harvard.edu/files/msen/files/race_causality.pdf)

- Walton and Cohen (2011), [A brief social-belonging intervention improves academic and health outcomes of minority students](https://www.marcellodibello.com/race-causality-discrimination/readings/Walton-BriefSocialBelongingIntervention-2011.pdf)

*Additional readings and materials*

- VanderWeele and Robinson (2014), [On Causal Interpretation of Race in Regressions Adjusting for Confounding and Mediating Variables](https://www.marcellodibello.com/race-causality-discrimination/readings/VanderWeeleRobinson-RaceRegression2024.pdf)



## 10/3


**Week 7: What is Race?**


We should switch gears and look at philosophical accounts of race. Admittedly, we cannot talk about whether race is a cause or how we can study the causality of race without asking, what is race? So we need to ask this question now. 
We'll start with a social constructionist view of race. 


*Essential readings*

- [Week 7 - handout](https://www.marcellodibello.com/race-causality-discrimination/handout/.html)  

- Haslanger social construction about race. Use the book “What is Race?” 

- Read the two essays with social constructionists views of race.

*Additional readings and materials*



## 10/10


**Week 8: Fall Break, No class**



## 10/17

**Week 9: What is Race?**



*Essential readings*

- [Week 10 - handout](https://www.marcellodibello.com/race-causality-discrimination/handout/.html)  



*Additional readings and materials*


## 10/24


**Week 10: Epistemic Injustice**

Epistemic bubbles and echo chambers are ecosystems of information, with their own internal rules and regulations. They constitute, in some sense, a common ground, at least for those inside them. A common ground, then, can also exclude, and perhaps it needs to exclude in order to include. So what happens when some people---along with their knowledge, evidence, testimony, experience---are excluded? Recently, a number of philosophers within the analytic tradition have theorized about "epistemic injustice"in various forms: testimonial, hermeneutical, oppressive. Long before these discussions in analytic philosophy, one of the foundational texts of post-colonial theory took up the question whether certain voices are systematically excluded from the conversation---and cannot even articulate their position. 

*Essential readings*

- [Week 10 - handout](https://www.marcellodibello.com/commonground/handout/epistemic-injustice)  

- Fricker (2007), Hermeneutical Injustice (Chapter 7 of [Epistemic Injustice](https://academic.oup.com/book/32817)) - *Check Canvas*

- Spivak (1988), [Can the Subaltern Speak?](https://abahlali.org/files/Can_the_subaltern_speak.pdf) (abridged version [here](https://jan.ucc.nau.edu/~sj6/Spivak%20CanTheSubalternSpeak.pdf))


*Additional readings and materials*

- Mills (2007), White Ignorance (Chapter 4 of [Black Rights/ White Wrongs
The Critique of Racial Liberalism](https://academic.oup.com/book/5106)) - *Check Canvas*

- Dotson (2014), [Conceptualizing Epistemic Oppression](https://www.marcellodibello.com/commonground/readings/DotsonEpistOppression.pdf)

- Spiegel (2022), [The Epistemic Injustice of Epistemic Injustice, Part I,](https://social-epistemology.com/2022/09/30/the-epistemic-injustice-of-epistemic-injustice-part-i-thomas-j-spiegel/)


## 3/23


**Week 11: Overlapping Consensus**

We now turn from the epistemological to the political. Citizens in liberal democracies often hold radically different views about the good life, religion, education, the common good. A question that has shaped political philosophy in the analytic tradition over the last three decades is, how can political deliberations and judicial decisions be carried out---and be legitimate---despite the fact of radical disagreement about matters of values among the citizenry? One answer is that, in the end, reasonable people will share a political common ground, an "overlapping consensus", which makes possible the exercise of public reason and its legitimacy. 

On this view, an important qualification must be made: an overlapping consensus must be sought between reasonable people. Those who are "unreasonable" do not enter the overlapping consensus, and thus are not entitled to pursue objections that may threaten the integrity and legitimacy of public reason. 

**Q&A with Jonathan Quong**

The first part of the class will consist in a Q&A session with [Jonathan Quong](https://dornsife.usc.edu/cf/phil/faculty_display.cfm?Person_ID=1048339), a moral and political philosopher at the University of Southern California whose work on unreasonable citizens is one of the readings assigned for today. *Please come prepared to ask questions.*


*Essential readings*

- [Week 11 - handout](https://www.marcellodibello.com/commonground/handout/political-liberalism.html)

- Rawls (1992), The Idea of an Overlapping Consensus and The Idea of Public Reason (lectures IV and VI of [Political Liberalism](http://cup.columbia.edu/book/political-liberalism/9780231130899)) - *Check Canvas*

- Quong (2010), Unreasonable Citizens (Chapter 10 of [Liberalism without Perfection](https://academic.oup.com/book/12006)) - *Check Canvas*

*Additional readings and materials*

- Cohen (2004), [Minimalism About Human Rights](https://fsi-live.s3.us-west-1.amazonaws.com/s3fs-public/evnts/media/j._cohen.minimalism.pdf)

<!---Earlier we discussed the topic of incommensurability from the point of view of language and conceptual schemes. Is there such a thing as value incommesurability? If so, can it be the explanation for the deep disagreements and ideological polarization that plague our political discourse today? As a preliminary step towards addressing this question, it is instructive to acquire a firm grasp of value incommensurability, its nature and scope. 
--->


<!---
- Broome (2000), [Incommensurable Values](https://philpapers.org/archive/BROIV-3.pdf)
*Additional readings and materials*
 - SEP, [Incommensurable Values](https://plato.stanford.edu/entries/value-incommensurable/)
--->


## 3/30

**Second essay due**


**Week 12: Self-organization**

The idea of an overlapping consensus presupposes that there is a minimal, essential set of values that we share (or procedures that we endorse), despite deep, widespread disagreement on many issues. But one wonders whether this overlapping consensus exists, and even if it does, one wonders whether it can ever be enough to give legitimacy to political deliberations. Another model suggests we need not worry about whether we have an overlapping consensus. A convergence might emerge, not as a matter of theoretical commitments to shared values, but as a self-organizing order of preferences, behaviors and practices. 


*Essential readings*

- [Week 12 - handout](https://www.marcellodibello.com/commonground/handout/self-org.html) 

 -  Gaus (2018), [Self-Organizing Moral Systems: Beyond Social Contract
Theory](http://www.gaus.biz/Gaus-Self-organizing.pdf)

- Gaus (2018), [The Complexity of a Diverse Moral Order](https://www.law.georgetown.edu/public-policy-journal/wp-content/uploads/sites/23/2018/11/16-S-Complexity-Diverse-Moral-Order.pdf) 

*Additional readings and materials*

- Troitzsch (2018), [Can Lawlike Rules Emerge without the Intervention of Legislators?](https://www.frontiersin.org/articles/10.3389/fsoc.2018.00002/full)


## 4/6


**Week 13: Individualism and conflict**

The ideal of an overlapping consensus (as well as the idea of a self-organizing convergence) must confront the reality of increasing conflicts, division, partisanship in our political lives. Perhaps, deep structural reasons are the cause, and these cannot be easily overcome by invoking a common ground. A couple of hypothesis are 
worth exploring here. First, an individualistic, competition-driven market logic that extends everywhere ends up pitting people against each another.  Second, political deliberations are inherently conflictual: friendship can exist only 
against the background of enmity.

*Essential readings*

- [Week 13 -handout](https://www.marcellodibello.com/commonground/handout/individualism.html) 

-  Hussain (2020), [Pitting People Against Each Other](https://www.researchgate.net/publication/339349346_Pitting_People_Against_Each_Other)

- 
Mouffe (1993), Politics and the Limits of Liberalism (Chapter 9 of [The Return of the Political](https://www.versobooks.com/books/3145-the-return-of-the-political)) - *Check Canvas*


*Additional readings and materials*


- Schmitt, The Concept of the Political (in [The Concept of the Political](https://press.uchicago.edu/ucp/books/book/chicago/C/bo5458073.html)) - *Check Canvas*

## 4/13

**Week 14: Trust and Civility**

If we still think that the ideal of an overlapping consensus and public reason are good ones---we might not---the question is how to regain a functioning public reason via targeted interventions. One proposal focuses on reestablishing trust, between people of opposing views and between citizens and public institutions. Another proposal emphasizes the virtue of civility. These proposals are not exclusive, and they both 
underscore that attitudes toward each other---not only an agreement about content---are essential for a functioning public reason. 

**Q&A with Cheshire Calhoun**

The first part of the class will consist in a Q&A session with [Cheshire Calhoun](https://en.wikipedia.org/wiki/Cheshire_Calhoun), a moral and political philosopher at Arizona State University whose work on the virtue of civility is one of the readings assigned for today. *Please come prepared to ask questions.*

*Essential readings*

- [Week 14 - handout](https://www.marcellodibello.com/commonground/handout/trust-civility.html) 

- Vallier (2019), Moral Peace and Social Trust (Chapter 1) and Public Justification (Chapter 3) of [Must Politics Be War?](https://www.kevinvallier.com/wp-content/uploads/2016/11/MPBW_v2.pdf)

- Calhoun (2000), [The Virtue Of Civility](https://www.marcellodibello.com/commonground/readings/CalhounCivility.pdf)


## 4/20

**Week 15: Workshop**


Students will workshop their research papers by giving short presentations accompanied by either a handout or slides and followed by Q&A. Each presenter will have 15 minutes to be split between the presentation itself and Q&A at the discretion of the presenter. The list of presenters is below:   

1:30 |  Angela  | title tba

1:45 | Andrew   | *Marked By An X: The Influence of Counterculture to A Better Common Ground*

2:00 | Cole  | *Common Agreement and Endoxa*

2:15 | Georgia   | title tba

2:30 | Kevin   | title tba

*15 minutes break*

3:00 | Nicholas | title tba

3:15 | Paola | *Ethics of Common Ground*

3:30 | Scout | *Contradictory Common Ground*

3:45 |  Vanessa | title tba

4:00 | Yashin | *Against Hermeneutical Justice: The Self-Sufficiency of Being*


## 4/27


**Third essay or research paper due**

**Week 16: Conclusion**

What have we learned about the common ground? 
What questions are still left open and 
what lines of inquiry seemed more promising?

- [Week 16 - handout](https://www.marcellodibello.com/commonground/handout/end.html) 


# Title IX 

Title IX is a federal law that provides that no person be excluded on the basis of sex from participation in, be denied benefits of, or be subjected to discrimination under any education program or activity.  Both Title IX and university policy make clear that sexual violence and harassment based on sex is prohibited.  An individual who believes they have been subjected to sexual violence or harassed on the basis of sex can seek support, including counseling and academic support, from the university.  If you or someone you know has been harassed on the basis of sex or sexually assaulted, you can find information and resources at https://sexualviolenceprevention.asu.edu/faqs. 

 

As a mandated reporter, I am obligated to report any information I become aware of regarding alleged acts of sexual discrimination, including sexual violence and dating violence.  ASU Counseling Services, https://eoss.asu.edu/counseling is available if you wish to discuss any concerns confidentially and privately. ASU online students may access 360 Life Services, https://goto.asuonline.asu.edu/success/online-resources.html.
